---
title: 'Blog 1: Tidy Models'
author: David Simbandumwe
subtitle: Value of Opinionated Workflows
output:
  word_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    theme: united
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
bibliography: references.bib
---





```{r}

library(tidyverse)
library(tidymodels)
library(workflowsets)
library(dotwhisker)
library(summarytools)

```


```{r}

#rm(list=ls())

cfg <- list(
              fileName = './inst/data/Teams.csv',
              rSource = './hw1/Functions.R',
              viewMethod = 'view' # options(render, browser, view)
          )


```




```{r}


knitr::opts_chunk$set(echo = F, 
                      warning = F, 
                      message = F, 
                      eval = T , 
                      results="asis", 
                      fig.height=6, 
                      fig.width=8)

set.seed(1234)


# styling
st_css()

 st_options(
   plain.ascii = FALSE,
   style = 'grid',
   dfSummary.style ='grid',
   freq.silent  = TRUE,
   headings     = FALSE,
   tmp.img.dir  = "./tmp",
   dfSummary.custom.1 =
     expression(
       paste(
         "Q1 - Q3 :",
         round(
           quantile(column_data, probs = .25, type = 2,
                    names = FALSE, na.rm = TRUE), digits = 1
         ), " - ",
         round(
           quantile(column_data, probs = .75, type = 2,
                    names = FALSE, na.rm = TRUE), digits = 1
         )
       )
     )
 )

#source('', local = knitr::knit_global())

```


# Overview




# Data Exploration

2.8  Teams table

yearID         Year
lgID           League
teamID         Team
franchID       Franchise (links to TeamsFranchise table)
divID          Team's division
Rank           Position in final standings
G              Games played
GHome          Games played at home
W              Wins
L              Losses
DivWin         Division Winner (Y or N)
WCWin          Wild Card Winner (Y or N)
LgWin          League Champion(Y or N)
WSWin          World Series Winner (Y or N)
R              Runs scored
AB             At bats
H              Hits by batters
2B             Doubles
3B             Triples
HR             Homeruns by batters
BB             Walks by batters
SO             Strikeouts by batters
SB             Stolen bases
CS             Caught stealing
HBP            Batters hit by pitch
SF             Sacrifice flies
RA             Opponents runs scored
ER             Earned runs allowed
ERA            Earned run average
CG             Complete games
SHO            Shutouts
SV             Saves
IPOuts         Outs Pitched (innings pitched x 3)
HA             Hits allowed
HRA            Homeruns allowed
BBA            Walks allowed
SOA            Strikeouts by pitchers
E              Errors
DP             Double Plays
FP             Fielding  percentage
name           Team's full name
park           Name of team's home ballpark
attendance     Home attendance total
BPF            Three-year park factor for batters
PPF            Three-year park factor for pitchers
teamIDBR       Team ID used by Baseball Reference website
teamIDlahman45 Team ID used in Lahman database version 4.5
teamIDretro    Team ID used by Retrosheet




```{r}

df <- read.csv(cfg$fileName)


# Preprocess Data
df <- df %>% select('yearID', 'lgID', 'teamID', 'franchID', 'name', 'divID', 'G', 'W', 'L', 
                    'R', 'H', 'X2B', 'X3B', 'HR', 
                    'RA', 'ER', 'HA', 'HRA')

df$wPer <- round(df$W / df$G, 3)
df$pythPer <- (df$R^2) / ((df$R^2) + (df$RA^2))


df <- df %>% mutate(era_cat = case_when(yearID >= 1969 ~ '1969+',
                                                yearID >= 1900 & yearID < 1969 ~ '1900-1969',
                                                yearID < 1900 ~ '1900-'))

df$era_cat <- factor(df$era_cat)



# Split Data
data_split <- initial_split(df, prop = 0.8, strata = era_cat)
train_data <- training(data_split)
test_data  <- testing(data_split)





```





```{r}

print(
  dfSummary(train_data %>% select(-c('teamID','franchID','name')), 
            varnumbers   = TRUE,
            na.col       = TRUE,
            graph.magnif = .8,
            tmp.img.dir  = "/tmp"),
  method = cfg$viewMethod
)


```




```{r}



```





```{r}


ggplot(sample_n(train_data, size=300, weight_by = era_cat),
       aes(x = wPer, 
           y = pythPer, 
           group = era_cat, 
           col = era_cat)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) +
  scale_color_viridis_d(option = "plasma", end = .7)

```


# Data Preparation

```{r}



```



```{r}



```



# Build Models 


```{r}

wins_rec <- 
  recipe(W ~ pythPer + era_cat + yearID + teamID + franchID , data = train_data)  %>%
  update_role(yearID, teamID, franchID, new_role = "ID") %>%
  step_dummy(all_nominal_predictors())


summary(wins_rec)


```



```{r}


win_lm_mod <- linear_reg() 


wins_wflow <- workflow() %>%
                  add_model(win_lm_mod) %>%
                  add_recipe(wins_rec)


```







```{r}

wins_fit <- 
  wins_wflow %>% 
  fit(data = train_data)


wins_fit %>% 
  extract_fit_parsnip() %>% 
  tidy()

```






# Predict

```{r}

predict(wins_fit, test_data)

```








# Model Set

```{r}
library(workflowsets)
```


```{r}


wins_list <- list(
  py = W ~ pythPer,
  hits = W ~ H,
  era = W ~ pythPer + era_cat
)

win_lm_mod <- linear_reg() 

win_models <- workflow_set(preproc = wins_list, models = list(lm = win_lm_mod))
win_models


win_models <-
   win_models %>%
   mutate(fit = map(info, ~ fit(.x$workflow[[1]], train_data)))
win_models




```




```{r}

win_models <-
   win_models %>%
   mutate(predict = map(fit, ~ predict(.x$workflow[[1]], test_data)))
win_models

```




```{r}

final_lm_res <- last_fit(lm_wflow, ames_split)
final_lm_res


```
























```{r}

lm_mod <- linear_reg() 

lm_fit <- 
  lm_mod %>% 
  fit(W ~ pythPer + era_cat, data = df)


```


```{r}

tidy(lm_fit) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))

```


## Predict New Values

```{r}

new_points <- expand.grid(pythPer = .5, 
                          era_cat = c("1900-", "1900-1969", "1969+"))
new_points



mean_pred <- predict(lm_fit, new_data = new_points)
mean_pred

```





```{r}
conf_int_pred <- predict(lm_fit, 
                         new_data = new_points, 
                         type = "conf_int")
conf_int_pred

# bind new columns
plot_data <- 
  new_points %>% 
  bind_cols(mean_pred) %>% 
  bind_cols(conf_int_pred)



# and plot:
ggplot(plot_data, aes(x = era_cat)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "wins")


```





```{r}



```


# Model Selection




# Conclusion









